### 2017-11-20 latex/mst_lukkarila_juri_2017.tex
edit

\university{Aalto University}
\school{School of Electrical Engineering}

\department{Department of Signal Processing and Acoustics}
\professorship{Speech recognition}

@@

%% English abstract.
\begin{abstractpage}[english]
Understanding and participating in conversations has been reported as one of the biggest challenges the hearing impaired face in their daily lives. These communication problems have been shown to have wide-ranging negative consequences, affecting the quality of life of these people and opportunities available to them in education and employment. A conversational assistance application design was investigated to alleviate these problems. The application uses automatic speech recognition technology to provide real-time speech-to-text transcriptions for the deaf and hard of hearing. To validate the method and investigate its usefulness, a prototype application was developed for use in user testing. User testing was performed with test participants representing the intended target user group. The results indicate that the Conversational Assistant approach is valid and helpful. Speech recognition accuracy, especially in noisy conditions, was identified as the primary target for further development for increased usefulness of the application. 
\end{abstractpage}

@@

\thesistitle{Keskusteluavustimen kehittäminen kuulovammaisia varten automaattista puheentunnistusta käyttäen}
\advisor{TkT Kalle Palomäki}
\department{Signaalinkäsittelyn ja akustiikan laitos}
\professorship{Puheentunnistus}
\keywords{puheentunnistus, ohjelmistokehitys, käyttäjäkeskeinen suunnittelu, käyttäjätestaus}
\begin{abstractpage}[finnish]
Keskustelupuheen ymmärtäminen ja keskusteluihin osallistuminen on raportoitu yhdeksi suurimmista haasteista, joita kuulovammaiset kohtaavat jokapäiväisessä elämässään. Näillä viestintäongelmilla on osoitettu olevan laaja-alaisia negatiivisia vaikutuksia, jotka heijastuvat näiden ihmisten elämänlaatuun ja heikentävät heidän yhdenvertaisia osallistumismahdollisuuksia opiskeluun ja työelämään. Työssä selvitettiin keskustelupuheelle suunnattua apusovellusta näiden ongelmien lievittämiseksi. Sovellus käyttää automaattista puheentunnistusta reaaliaikaiseen puheen tekstittämiseen kuurojen ja huonokuuloisien tukemista varten. Menetelmän validoimiseksi ja sen hyödyllisyyden tutkimiseksi kehitettiin prototyyppisovellus käyttäjätestaukseen. Käyttäjäkoe suoritettiin sovelluksen kohderyhmää edustavilla koekäyttäjillä. Saadut tulokset viittaavat siihen, että esitetty Keskusteluavustin on toimiva ja hyödyllinen apuväline. Puheentunnistustarkkuus erityisesti meluisissa olosuhteissa osoittautui ensisijaiseksi kehityskohteeksi apusovelluksen hyödyllisyyden lisäämiseksi. 
\end{abstractpage}

@@

This work was carried out during the spring and summer of 2017 at the Speech Recognition research group of the Department of Signal Processing and Acoustics at Aalto University School of Electrical Engineering. This thesis would not have been possible without the support of the Academy of Finland for the project \textit{Conversation Assistant for the Hearing Impaired}. \\\\
First and foremost, I would like to thank supervisor Prof.\ Mikko Kurimo and my advisor D.Sc.\ Kalle Palomäki for giving me the opportunity to work on this interesting and multifaceted topic. I am grateful for the freedom and continued support given to me during this work. I would like to thank Prof.\ Ville Pulkki, Symeon Delikaris-Manias and Juhani Paasonen from the Aalto Spatial Sound research group for assistance with the background noise recordings used in the user tests, and Ilkka Huhtakallio for providing the necessary audio equipment. My thanks to Olli Savisaari from the User Interfaces research group for consulting with user testing methods. From the Speech Recognition group, I would like to thank Seppo Enarvi, Reima Karhila, Katri Leino, Ulpu Remes, Aku Rouhe and Peter Smit for general help, tips and discussion. \\\\
I would like to thank Päivi Rainò from HUMAK and Tarja Kaikkonen from Kuuloliitto ry for helping to recruit deaf and hard of hearing persons for the user tests, and many thanks to the people who participated in the tests. \\\\
Finally, I would like to thank Siiri for all the support and motivation. \\

@@

\vspace{5mm}
\noindent Juri Akseli Lukkarila

@@

{\setlength{\tabcolsep}{6mm}
\begin{tabular}{ll}
ASR         	 & Automatic speech recognition \\
AVSR			& Audio-visual speech recognition \\
DCT				& Discrete cosine transformation \\
DNN				& Deep neural network \\
DSP				& Digital signal processing \\
FST				& Finite-state transducer \\
GPGPU 		  	& General-purpose computing on graphics processing units \\
GUI      		  & Graphical user interface \\
HCI  			  & Human-computer interaction \\
HMM   			& Hidden Markov model \\
LER				 & Letter error rate \\
LVCSR 		   & Large vocabulary continuous speech recognition \\
MFCC  		   & Mel-frequency cepstral coefficient \\
OS				& Operating system \\
PCM  		    & Pulse-code modulation \\
RNN				& Recurrent neural network \\
SNR				 & Signal-to-noise ratio \\
SPL				 & Sound pressure level \\
STT					& Speech-to-text \\
UI	       		   & User interface \\
WER    			& Word error rate \\
WFST		  & Weighted finite-state transducer
\end{tabular}}

@@

Understanding and participating in conversations has been reported as one of the most prominent problems that hearing impaired individuals face in their daily lives by a wide variety of studies and reports, ranging from medical research and engineering to sociology and beyond \cite{moore2007cochlear, peterson2010cochlear, wilson2017global, ohlenforst2017effects, stacey2006hearing, healy2016difficulty, hietala2008huonokuuloinen, koskela2013kuulokojeen, lavikainen2014, blomberg2012sisakorvaistutetta, haatainen2013viestintahaasteet}. Experiences from previous projects at Aalto University's Speech Recognition group also point to the same conclusion. Conversations, and speech in general, form a major part of all human interaction, and the reduced or completely lost capability for conversations due to hearing impairment can have wide-ranging consequences not only on the hearing impaired individual directly, but also on the whole of society through social and economic factors \cite{koskela2013kuulokojeen, stacey2006hearing}. \\\\
Losing this substantial part of social interaction can have a strong negative effect on a person quality of life and affect the opportunities available to them, for instance in education and employment \cite{hietala2008huonokuuloinen, lavikainen2014}. Therefore, it is ultimately a matter of equality and equal opportunities for the deaf and hard of hearing members of society. Furthermore, statistical studies on the prevalence of hearing impairment have reported that the burden of hearing loss has been constantly increasing around the world, and is presently higher than ever before \cite{mathers2000global, wilson2017global}. An estimated 500 million people around the world had disabling hearing loss in 2015, which translates to approximately seven percent of the world’s population at that time \cite{wilson2017global}. New medical and technological solutions have been called for in order to effectively treat hearing impairment, as the number of afflicted people keeps growing worldwide \cite{wilson2017global}. \\\\
Currently, no medical cure exists for the majority of hearing loss cases, in the sense that the normal biological operation of the ear is restored \cite{moore2007cochlear}. Sensorineural hearing loss is by far the most common type of hearing loss, and damage or abnormalities in the hair cells of the ear are a typical cause for it \cite{moore2007cochlear, koskela2013kuulokojeen}. Hair cells, which are responsible for translating the mechanical vibration of sound to electrochemical signals for the brain, do not regrow naturally, and once damaged, cannot be repaired with currently available medical methods \cite{moore2007cochlear}. Consequently, when a hair cell is damaged, the resulting loss in hearing is physiologically permanent, though modern treatments like the cochlear implant can partially restore hearing sensations \cite{moore2007cochlear}. Damage to hair cells is most commonly caused by exposure to excessive noise and through time by the aging process. Gene therapy and stem cell--based methods are being researched as a potential solution, and might someday enable the regeneration of hair cells  \cite{mclean2017clonal}, but at least for the foreseeable future no tangible cure is available. Instead, existing medical treatments rely on augmenting and amplifying the degree of hearing still present with personal electronic hearing devices, or in the case of severe hearing loss and deafness, through the surgically implanted cochlear implant that bypasses the outer and middle ear altogether \cite{moore2007cochlear}. In addition to the electrical devices focused on improving the level of an affected individuals hearing, spoken communication between persons is largely supported by translating sound to text and by using sign language, with both methods relying heavily on human translators \cite{moore2007cochlear, raino2012sisakorvaistutteen}.

@@

\textit{Automatic speech recognition} (ASR) is the science and technology of transcribing spoken language into written words automatically using computers \cite{yu2014automatic, huang2001spoken}. The history of speech recognition research goes as far back as the 1950s, when the first steps were taken at the renowned Bell Laboratories \cite{gales2008application}. For a long time, working applications were comprised mostly of simple voice user interfaces and dictation systems with limited, application-specific vocabularies \cite{yu2014automatic, gales2008application}. The holy grail of speech recognition technology has arguably been real-time, speaker independent recognition of unlimited vocabularies, of which everyday conversational speech is a good example. In recent years, significant advances have been made especially in this area, referred to in ASR research as \textit{large vocabulary continuous speech recognition} (LVCSR) \cite{yu2014automatic, keronen2014approaching}. These advances have been enabled in large part thanks to the fast development and adoption of new machine learning techniques \cite{yu2014automatic, hinton2012deep}. In particular, \textit{deep neural networks} (DNN) with multiple hidden layers (hence the word "deep") have proven to be of great value for ASR systems \cite{yu2014automatic, hinton2012deep}. Machine learning itself has become widespread and practical owing to the advances in computer hardware and computational resources available, especially through the popularization of the so called \textit{general-purpose computing on graphics processing units} (GPGPU) technology for massively parallel computation, that has enabled efficient training of neural networks with relatively inexpensive and widely available computing hardware \cite{yu2014automatic, hinton2012deep}. \\\\
As a consequence, automatic speech recognition has gained a great deal of popularity during the last few years, and is starting to be widely used in practice in the commercial landscape and everyday technical applications, from smartphones and tablets to personal computers \cite{yu2014automatic}. Today, companies actively promoting their own speech recognition systems to average consumers include influential giants like Amazon, Apple, Google and Microsoft. Services based on speech recognition technology encompasses voice-controlled personal assistants, dictation, automatic captioning, voice-based search of audio and video content, and a wide variety of voice user interfaces \cite{yu2014automatic, li2014overview}. While there are still some specific, and in many cases, quite substantial challenges for automatic speech recognition, in general it has become accurate and reliable enough to be used in many practical applications requiring continuous recognition of large vocabularies \cite{yu2014automatic, keronen2014approaching, mcgraw2016personalized}. This is demonstrated well by the speech recognition based services of the previously mentioned large technology companies, such as Apple's Siri voice assistant \cite{li2014overview}. At the same time, mobile smart devices (phones, tablets and laptops) and  wireless internet access have become ubiquitous in all developed countries of the world, offering a convenient platform for utilizing speech recognition technology in practically every conceivable place and situation \cite{yu2014automatic, mcgraw2016personalized}.

@@

Returning to the communication problems hearing impaired people encounter, having automatic real-time transcriptions of speech always available could potentially be extremely helpful in many of these problematic situations. Realizing this type of system offers a challenging, but well-defined practical application of modern ASR technology to a concrete problem: The goal of this work was to develop and test a conversational assistance application aimed for deaf and hard of hearing individuals, which translates speech into text in real time using automatic speech recognition. The intended purpose of this assistive application, henceforth referred to as the \textit{Conversation Assistant}, is to help and support hearing impaired persons in conversations, and other situations where they are being spoken to, such as meetings, lectures and school classes. It is not intended to fully replace other personal assistive devices like hearing aids, but instead to supplement them. \\\\
The basic operation principle of the proposed Conversation Assistant is illustrated in figure \ref{fig:assistant}, describing a typical conversational scenario where the Conversation Assistant could be used. 
\begin{figure}[b]
	\centering 
	\includegraphics[trim={0cm 0cm 0cm 0cm}, clip, width=\textwidth]{assistant3.pdf}
	\caption{The operation principle and an example use case for the Conversation Assistant. In the picture, the person on the left is speaking. The person on the right uses the Conversation Assistant, which converts that speech into text in real-time.}
	\label{fig:assistant} 
\end{figure} 
Ideally, the Conversation Assistant could be used with any applicable smart device the user already owns, as long as the basic requirements are met. The Conversation Assistant approach is described in detail in section \ref{sec:implementation}. Using automatic speech recognition technology to support deaf and hard of hearing individuals in some capacity has been proposed and tested already previously \cite{robison1996computer, kheir2007inclusion, mirzaei2012combining}. However, many of these studies have focused only on some particular setting or situation, such as school classrooms, or have been otherwise limited in their scope. Also, automatic speech recognition technology has matured considerably during this decade, meaning that the findings of many earlier studies might not be accurate any longer. Furthermore, to the extend of our knowledge, this type of system has not been investigated or implemented before for the Finnish language.

@@

Some deaf and hard of hearing persons use sign language as an alternative for spoken communication. While this can work well for people who are fluent in it, the problem is that very few people know sign language, especially outside the deaf community. In fact, the share of sign language users among the hearing impaired has been constantly decreasing as a result of modern treatment technology \cite{stacey2006hearing, raino2012sisakorvaistutteen}. In particular, the cochlear implant has enabled a significant share of prelingually\footnote{before language acquisition, including congenital cases (present at birth).} deaf children to not require sign language for communication anymore \cite{moore2007cochlear, peterson2010cochlear}. Different studies estimate that in the near future, approximately 60-80\% of prelingually deaf children will use speech as their main means of communication, as opposed to sign language, thanks to the improved auditory perception provided by cochlear implants \cite{raino2012sisakorvaistutteen}. As a consequence, the need and incentive for the general public to learn sign language in order to communicate with hearing impaired individuals is diminishing even further. As the technological and medical solutions continue to advance, sign language is slowly becoming obsolete. Indeed, sign language is feared and predicted to become extinct in the coming decades \cite{raino2012sisakorvaistutteen}. \\\\
While modern hearing aids and cochlear implants have become very sophisticated by utilizing digital technology and signal processing, there remains many challenges and limitations in their everyday use \cite{levitt2007historical}. One of the major obstacles can be the cost: High-end, personalized devices and surgery are expensive, especially if not covered by health insurance or provided by a public healthcare system \cite{wilson2017global}. Adherence to hearing aid use and rehabilitation can be poor: It has been estimated that only 20-50\% of people who would benefit from a hearing aid are actually using one \cite{koskela2013kuulokojeen}. Even with the extensive digital signal processing and noise reduction in current devices, speech intelligibility in situations with background noise remains one of the major challenges \cite{healy2016difficulty, levitt2007historical, goehring2016speech}. Cochlear implants in particular appear to be very susceptible to noise with a dramatic reduction in speech perception quality in noisy conditions \cite{ healy2016difficulty, friesen2001speech, fu2005noise, srinivasan2013improving}. For public spaces and events, the effects of noise and the environment can be alleviated with a specially installed induction loop, commonly referred to as a \textit{hearing loop} \cite{salonen2013hearing}: The desired sound source is fed electrically into a current-carrying wire loop, and the resulting electromagnetic field containing the baseband audio signal is picked up directly by a hearing aid or other device. The pickup coil in a hearing aid or implant is commonly referred to as a \textit{telecoil} (or T-coil). Typical installation locations include airports, auditoriums, concert halls and public bureau buildings. FM systems are a similar alternative for induction loops, which use radio transmission instead of electromagnetic induction. Naturally, these solutions are also not without some technical and practical complications. Interference from metallic structures and other equipment can be an issue, leading to an uneven field strength and affecting the reception quality. One very concrete problem is that many places don't have them installed yet \cite{healy2016difficulty, wilson2017global}. \\\\
Currently, human sign and written language interpreters have a large role in facilitating face-to-face communication between hearing and non-hearing persons, especially in more formal situations that can be scheduled in advance \cite{raino2012sisakorvaistutteen, pereira2010communication, gaur2016effects}. Written language interpreters translate speech to text simultaneously with a speaker by manually typing it into a computer. Written language interpretation happens only in one direction, whereas sign language interpretation can be bidirectional: first, the sign language interpreter translates speech into signs. Then the sign language user can respond with signs, which are then spoken aloud by the translator. Interpreters are used for example in classrooms, meetings, public events and private appointments. Requiring such an extra person for communication has many obvious disadvantages \cite{gaur2016effects}. Firstly, there are the multitude of practical challenges: Interpreters are not available at all times and in all situations. Their number and availability can be quite limited especially outside urban population centers. Professional interpreters typically require a multi-year education and training, limiting their number and introducing costs. Privacy and self-determination can also be a concern, even though interpreters are customarily bound by confidentiality \cite{pereira2010communication}. \\\\
In conclusion, all the existing traditional solutions have their own problems. One of the fundamental issues with many of these are the costs associated with them, both for the individual and for society \cite{wilson2017global}. Our proposed, automatic speech recognition based solution has the potential to be a relatively inexpensive and highly cost-efficient alternative, enabling communication for the hearing impaired in most everyday situations quickly and conveniently. Ideally, the Conversation Assistant could also remove the need for human interpreters in many situations, or at least function as a workable alternative when human interpreters are not present or available.

@@

Hearing impairment can be defined as having a reduced or deficient hearing ability, generally caused as a result of decreased hearing sensitivity in one or both ears \cite{moore2007cochlear, ohlenforst2017effects}. Neurological conditions affecting auditory processing in the brain have also been identified, though they are quite rare and hard to diagnose. These \textit{auditory processing disorders} show as various difficulties in recognizing and interpreting sounds correctly, even though the ears are functioning physiologically normally \cite{deafness}. The terms \textit{hearing impairment} and \textit{hearing loss} ar commonly used synonymously \cite{moore2007cochlear}. In this work, \textit{hearing loss} is used to refer specifically to the sensory impairment of hearing, and \textit{hearing impairment} to the overall condition and disability resulting from auditory dysfunction. \\\\

>>>>

\university{Aalto University}

\school{School of Electrical Engineering}

\department{Department of Signal Processing and Acoustics}

\professorship{Speech and Language Processing}

@@

%% English abstract.
\begin{abstractpage}[english]
Understanding and participating in conversations has been reported as one of the biggest challenges hearing impaired people face in their daily lives. These communication problems have been shown to have wide-ranging negative consequences, affecting their quality of life and the opportunities available to them in education and employment. \\

A conversational assistance application was investigated to alleviate these problems. The application uses automatic speech recognition technology to provide real-time speech-to-text transcriptions to the user, with the goal of helping deaf and hard of hearing persons in conversational situations. To validate the method and investigate its usefulness, a prototype application was developed for testing purposes using open-source software. A user test was designed and performed with test participants representing the target user group. \\

The results indicate that the Conversation Assistant method is valid, meaning it can help the hearing impaired to follow and participate in conversational situations. Speech recognition accuracy, especially in noisy environments, was identified as the primary target for further development for increased usefulness of the application. Conversely, recognition speed was deemed to be sufficient and already surpass the transcription speed of human transcribers.
\end{abstractpage}

@@

\thesistitle{Keskusteluavustimen kehittäminen kuulovammaisia varten automaattista puheentunnistusta käyttäen}
\advisor{TkT Kalle Palomäki}
\department{Signaalinkäsittelyn ja akustiikan laitos}
\professorship{Puheen- ja kielenkäsittely}
\keywords{puheentunnistus, ohjelmistokehitys, käyttäjäkeskeinen suunnittelu, käyttäjätestaus}
\begin{abstractpage}[finnish]
Keskustelupuheen ymmärtäminen ja keskusteluihin osallistuminen on raportoitu yhdeksi suurimmista haasteista, joita kuulovammaiset kohtaavat jokapäiväisessä elämässään. Näillä viestintäongelmilla on osoitettu olevan laaja-alaisia negatiivisia vaikutuksia, jotka heijastuvat elämänlaatuun ja heikentävät kuulovammaisten yhdenvertaisia osallistumismahdollisuuksia opiskeluun ja työelämään. \\

Työssä kehitettiin ja arvioitiin apusovellusta keskustelupuheen ymmärtämisen ja keskusteluihin osallistumisen helpottamiseksi. Sovellus käyttää automaattista puheentunnistusta reaaliaikaiseen puheen tekstittämiseen kuuroja ja huonokuuloisia varten. Menetelmän toimivuuden vahvistamiseksi ja sen hyödyllisyyden tutkimiseksi siitä kehitettiin prototyyppisovellus käyttäjätestausta varten avointa lähdekoodia hyödyntäen. Testaamista varten suunniteltiin ja toteutettiin käyttäjäkoe sovelluksen kohderyhmää edustavilla koekäyttäjillä. \\

Saadut tulokset viittaavat siihen, että työssä esitetty Keskusteluavustin on toimiva ja hyödyllinen apuväline huonokuuloisille ja kuuroille. Puheentunnistustarkkuus erityisesti meluisissa olosuhteissa osoittautui ensisijaiseksi kehityskohteeksi apusovelluksen hyödyllisyyden lisäämiseksi. Puheentunnistuksen nopeus arvioitiin puolestaan jo riittävän nopeaksi, ylittäen selkeästi kirjoitustulkkien kirjoitusnopeuden. 
\end{abstractpage}

@@

This work was carried out during the spring and summer of 2017 at the Speech Recognition research group of the Department of Signal Processing and Acoustics at the Aalto University School of Electrical Engineering. This thesis would not have been possible without the support of the Academy of Finland for the project \textit{Conversation Assistant for the Hearing Impaired}. \\\\
First and foremost, I would like to thank supervisor Prof.\ Mikko Kurimo and my advisor D.Sc.\ Kalle Palomäki for giving me the opportunity to work on this interesting and multifaceted topic. I am grateful for the freedom and continued support given to me during this work. I would like to thank Prof.\ Ville Pulkki, Symeon Delikaris-Manias and Juhani Paasonen from the Aalto Spatial Sound research group for assistance with the background noise recordings used in the user tests, and Ilkka Huhtakallio for providing the necessary audio equipment. My thanks to Olli Savisaari from the User Interfaces research group for consulting with user testing methods. From the Speech Recognition group, I would like to thank Seppo Enarvi, Reima Karhila, Katri Leino, Ulpu Remes, Aku Rouhe and Peter Smit for general help, tips and discussion. \\\\
I would like to thank Päivi Rainò from HUMAK and Tarja Kaikkonen from Kuuloliitto ry for helping to recruit deaf and hard of hearing persons for the user tests, and many thanks to the people who participated in the tests. \\\\
Finally, I would like to thank Siiri for all the support and motivation. \\

@@

\vspace{2mm}
\noindent Juri Akseli Lukkarila

@@

{\setlength{\tabcolsep}{6mm}
\begin{tabular}{ll}
ASR         	 & Automatic speech recognition \\
AVSR			& Audio-visual speech recognition \\
DNN				& Deep neural network \\
DSP				& Digital signal processing \\
FST				& Finite-state transducer \\
GPGPU 		  	& General-purpose computing on graphics processing units \\
GUI      		  & Graphical user interface \\
HCI  			  & Human-computer interaction \\
HMM   			& Hidden Markov model \\
LER				 & Letter error rate \\
LVCSR 		   & Large vocabulary continuous speech recognition \\
MFCC  		   & Mel-frequency cepstral coefficient \\
OS				& Operating system \\
PCM  		    & Pulse-code modulation \\
RNN				& Recurrent neural network \\
SNR				 & Signal-to-noise ratio \\
SPL				 & Sound pressure level \\
STT					& Speech-to-text \\
UI	       		   & User interface \\
WER    			& Word error rate \\
WFST		  & Weighted finite-state transducer
\end{tabular}}

\vspace{5mm}

@@

Understanding and participating in conversations has been reported as one of the most prominent problems that hearing impaired individuals face in their daily lives by a wide variety of studies and reports, ranging from medical research and engineering to sociology and beyond \cite{moore2007cochlear, peterson2010cochlear, wilson2017global, ohlenforst2017effects, stacey2006hearing, healy2016difficulty, hietala2008huonokuuloinen, koskela2013kuulokojeen, lavikainen2014, blomberg2012sisakorvaistutetta, haatainen2013viestintahaasteet}. Experiences from previous projects at the Aalto University Speech Recognition group also point to the same conclusion. Conversations, and speech in general, form a major part of all human interaction, and the reduced or completely lost capability for conversations due to hearing impairment can have wide-ranging consequences not only on the hearing impaired individual directly, but also on the whole of society through social and economic factors \cite{stacey2006hearing, koskela2013kuulokojeen}. Losing this substantial part of social interaction can have a strong negative effect on a person quality of life, and affect the opportunities available to them for instance in education and employment \cite{hietala2008huonokuuloinen, lavikainen2014}. Therefore, it is ultimately a matter of equality and equal opportunities for the deaf and hard of hearing members of society. Furthermore, statistical studies on the prevalence of hearing impairment have reported that the burden of hearing loss has been constantly increasing around the world, and is presently higher than ever before \cite{wilson2017global, mathers2000global}. An estimated 500 million people around the world had disabling hearing loss in 2015, which translates to approximately seven percent of the world’s population at that time \cite{wilson2017global}. New medical and technological solutions have been called for in order to effectively treat hearing impairment, as the number of afflicted people keeps growing worldwide \cite{wilson2017global}. \\\\
Currently, no medical cure exists for the majority of hearing loss cases, in the sense that the normal biological operation of the ear is restored \cite{moore2007cochlear}. Sensorineural hearing loss is by far the most common type of hearing loss, and damage or abnormalities in the hair cells of the ear are a typical cause for it \cite{moore2007cochlear, koskela2013kuulokojeen}. Hair cells, which are responsible for translating the mechanical vibration of sound to electrochemical signals for the brain, do not regrow naturally, and once damaged, cannot be repaired with currently available medical methods \cite{moore2007cochlear}. Consequently, when a hair cell is damaged, the resulting loss in hearing is physiologically permanent, though modern treatments like the cochlear implant can partially restore hearing sensations \cite{moore2007cochlear, peterson2010cochlear}. Damage to hair cells is most commonly caused by exposure to excessive noise and through time by the aging process. However, gene therapy and stem cell--based methods are being researched as a potential solution, and might someday enable the regeneration and restoration of hair cells  \cite{mclean2017clonal}. Existing medical treatments rely on augmenting and amplifying the degree of hearing still present with personal electronic hearing devices, or in the case of severe hearing loss and deafness, through a surgically implanted cochlear implant that bypasses the outer and middle ear altogether \cite{moore2007cochlear}. In addition to the electrical devices focused on improving the level of an affected individuals hearing, spoken communication between persons is largely supported by translating sound to text and by using sign language, with both methods relying heavily on human translators \cite{moore2007cochlear, raino2012sisakorvaistutteen}.

@@

\textit{Automatic speech recognition} (ASR) is the science and technology of transcribing spoken language into written words automatically using computers \cite{yu2014automatic, huang2001spoken}. The history of speech recognition research goes as far back as the 1950s, when the first steps were taken at the renowned Bell Laboratories \cite{gales2008application}. For a long time, working applications were comprised mostly of simple voice user interfaces and dictation systems with limited, application-specific vocabularies \cite{yu2014automatic, gales2008application}. The holy grail of speech recognition technology has arguably been real-time, speaker independent recognition of unlimited vocabularies, of which everyday conversational speech is a good example. In recent years, significant advances have been made especially in this area, referred to in ASR research as \textit{large vocabulary continuous speech recognition} (LVCSR) \cite{yu2014automatic, keronen2014approaching}. These advances have been enabled in large part thanks to the fast development and adoption of new machine learning techniques \cite{yu2014automatic, hinton2012deep}. In particular, \textit{deep neural networks} (DNN) with multiple hidden layers (hence the word \textit{"deep"}) have proven to be of great value for ASR systems \cite{yu2014automatic, hinton2012deep}. Machine learning itself has become widespread and practical owing to the advances in computer hardware and computational resources available, especially through the popularization of the so called \textit{general-purpose computing on graphics processing units} (GPGPU) technology for massively parallel computation, which has enabled efficient training of neural networks with relatively inexpensive and widely available computing hardware \cite{yu2014automatic, hinton2012deep}. \\\\
As a consequence, automatic speech recognition has gained a great deal of popularity during the last few years, and is becoming widely used in practice in the commercial landscape and everyday applications from smartphones and tablets to personal computers \cite{yu2014automatic}. Today, companies actively promoting their own speech recognition systems to average consumers include influential giants like Amazon, Apple, Google and Microsoft \cite{xiong2017microsoft}. Services based on speech recognition technology encompasses voice-controlled personal assistants, dictation, automatic captioning, voice-based search of audio and video content, and a wide variety of voice user interfaces \cite{yu2014automatic, li2014overview}. While there are still some specific, and in many cases, quite substantial challenges for automatic speech recognition, in general it has become accurate and reliable enough to be used in many practical applications requiring continuous recognition of large vocabularies \cite{yu2014automatic, keronen2014approaching, mcgraw2016personalized}. This is demonstrated well by the speech recognition based services of the previously mentioned large technology companies, such as Apple's Siri voice assistant \cite{li2014overview}. At the same time, mobile smart devices such as phones, tablets and laptops, in combination with fast wireless internet access have become ubiquitous in all developed countries of the world, offering a convenient platform for utilizing speech recognition technology practically in all places and situations \cite{yu2014automatic, mcgraw2016personalized}.

@@

Returning to the spoken communication problems hearing impaired people encounter, having automatic real-time transcriptions of speech always available could potentially be extremely helpful in many of these problematic situations. Realizing this type of system offers a challenging, but well-defined practical application of modern speech recognition technology to a concrete problem. The goal of this work was to develop and test a conversational assistance application aimed for deaf and hard of hearing individuals, which translates speech into text in real time using automatic speech recognition. The intended purpose of this assistive application, henceforth referred to as the \textit{Conversation Assistant}, is to help and support hearing impaired persons in conversations and other situations where they are being spoken to, such as meetings, lectures and school classrooms. It is not intended to fully replace other personal assistive devices like hearing aids, but instead to supplement them. \\\\
The basic operation principle of the proposed Conversation Assistant is illustrated in figure \ref{fig:assistant}, which presents a typical conversational scenario where the Conversation Assistant could be used.
\begin{figure}[b]
	\centering 
	\includegraphics[trim={0cm 0cm 0cm 0cm}, clip, width=\textwidth]{assistant3.pdf}
	\caption{The operation principle and an example use case for the Conversation Assistant. In the illustration, the person on the left is speaking. The person on the right uses the Conversation Assistant, which converts the speech into text in real-time.}
	\label{fig:assistant} 
\end{figure} 
Ideally, the Conversation Assistant could be used with any applicable smart device the user already owns, as long as the basic requirements are met. The Conversation Assistant method is described in detail in section \ref{sec:implementation}. Using automatic speech recognition technology to support deaf and hard of hearing individuals in some form or capacity has been proposed and tested already previously \cite{robison1996computer, kheir2007inclusion, mirzaei2012combining}. However, many of these studies have focused only on some particular setting or situation, such as school classrooms, or have otherwise been limited in their scope. Also, automatic speech recognition technology has matured considerably during this decade, meaning that the findings of many earlier studies might not be accurate any longer. A similar communication aid using Finnish speech recognition was investigated by Karjalainen et al. already in 1997. However, to the extend of our knowledge, no implementations of an application like the Conversation Assistant are currently available for the Finnish language.

@@

Some deaf and hard of hearing persons use sign language as an alternative for spoken communication. While this can work well for people who are fluent in it, the problem is that very few people know sign language, especially outside the deaf community. In fact, the share of sign language users among the hearing impaired has been constantly decreasing as a result of modern treatment technology \cite{stacey2006hearing, raino2012sisakorvaistutteen}. In particular, the cochlear implant has enabled a significant share of prelingually\footnote{before language acquisition, including congenital cases (present at birth).} deaf children not to require sign language for communication anymore \cite{moore2007cochlear, peterson2010cochlear}. Different studies estimate that in the near future, approximately 60-80\% of prelingually deaf children will use speech as their primary means of communication, as opposed to sign language, thanks to the improved auditory perception provided by cochlear implants \cite{raino2012sisakorvaistutteen}. As a consequence, the need and incentive for the general public to learn sign language in order to communicate with hearing impaired individuals is diminishing even further. As the technological and medical solutions continue to advance, sign language is slowly becoming obsolete. Indeed, sign language is though to be in danger of becoming extinct in the coming decades \cite{raino2012sisakorvaistutteen}. \\\\
While modern hearing aids and cochlear implants have become very sophisticated by utilizing digital technology and signal processing, there remains many challenges and limitations in their everyday use \cite{levitt2007historical}. One of the major obstacles can be the cost: High-end, personalized devices and surgery are expensive, especially if not covered by health insurance or provided by a public healthcare system \cite{wilson2017global}. Adherence to hearing aid use and rehabilitation can be low: It has been estimated that only 20-50\% of the people who would benefit from a hearing aid are actually using one \cite{koskela2013kuulokojeen}. Even with the extensive digital signal processing and noise reduction in current devices, speech intelligibility in situations with background noise remains one of the major challenges \cite{healy2016difficulty, levitt2007historical, goehring2016speech}. Cochlear implants in particular appear to be very susceptible to noise with a dramatic reduction in speech perception quality in noisy conditions \cite{ healy2016difficulty, friesen2001speech, fu2005noise, srinivasan2013improving}. For public spaces and events, the effects of noise and the environment can be alleviated with a specially installed induction loop, commonly referred to as a \textit{hearing loop} \cite{salonen2013hearing}: The desired sound source is fed electrically into a current-carrying wire loop, and the resulting electromagnetic field containing the baseband audio signal is picked up directly by a hearing aid or other device. The pickup coil in a hearing aid or implant is commonly referred to as the \textit{telecoil} (or \textit{T-coil}). Typical installation locations include airports, auditoriums, concert halls and public bureau buildings. FM systems are a similar alternative for induction loops that use radio transmission instead of electromagnetic induction. Naturally, these solutions are not without some technical and practical complications. Interference from metallic structures and other equipment can be an issue, leading to an uneven field strength and affecting the reception quality. One very concrete problem is that many places don't have them installed (yet) \cite{wilson2017global, healy2016difficulty}. \\\\
Currently, human sign and written language interpreters have a large role in facilitating face-to-face communication between hearing and non-hearing persons, especially in more formal situations that can be scheduled in advance \cite{raino2012sisakorvaistutteen, pereira2010communication, gaur2016effects}. Written language interpreters translate speech to text simultaneously with a speaker by manually typing it into a computer. Written language interpretation happens only in one direction, whereas sign language interpretation can be bidirectional: first, the sign language interpreter translates speech into signs. Then the sign language user can respond with signs, which are then spoken aloud by the translator. Interpreters are used for example in classrooms, meetings, public events and private appointments. Requiring an extra person for communication has many obvious disadvantages \cite{gaur2016effects}. Firstly, there are the multitude of practical challenges, like that interpreters are not available at all times and in all situations. Their number and availability can be quite limited especially outside urban population centers. Professional interpreters typically require a multi-year education and training, limiting their number and introducing costs. Interpretation can lead to a reduced possibility for self-determination \cite{pereira2010communication}. Privacy and communication of sensitive matters can also be a concern, even though interpreters are customarily bound by confidentiality. \\\\
In conclusion, all the existing traditional solutions have their own problems. One of the fundamental issues with many of the above-mentioned are the costs associated with them, both for the individual and for society \cite{wilson2017global}. The proposed ASR-based solution has the potential to be a relatively inexpensive and highly cost-efficient alternative, enabling communication for the hearing impaired in everyday conversational situations quickly and conveniently. Ideally, the Conversation Assistant could also remove the need for human interpreters in many situations, or at least function as a workable alternative when human interpreters are not available.

@@

Hearing impairment can be defined as having a reduced or deficient hearing ability, generally caused as a result of decreased hearing sensitivity in one or both ears \cite{moore2007cochlear, ohlenforst2017effects}. Neurological conditions affecting auditory processing in the brain have also been identified, though they are quite rare and hard to diagnose. These \textit{auditory processing disorders} show as various difficulties in recognizing and interpreting sounds correctly, even though the ears are functioning physiologically normally \cite{deafness}. The terms \textit{hearing impairment} and \textit{hearing loss} ar commonly used synonymously \cite{moore2007cochlear}. In this work, \textit{hearing loss} is used to refer specifically to sensory impairment of hearing, and \textit{hearing impairment} to the overall condition and disability resulting from auditory dysfunctions. \\\\
